\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-09 k-planar.tex
Vida's changes and other stuff

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs (and other more general classes). Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }

The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to the setting of $k$-planar graphs (and other more general classes).

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
So, loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{kPlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).

@@

We actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$, and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

\cref{ShortcutProduct,AddDummy} immediately imply \cref{kPlanarBasic} with $O(k^{11})$ instead of $O(k^5)$. Some further observations reduce the bound to $O(k^5)$; see \cref{Tools}.

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$. If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

Note that the dependence on $k$ in the bound on the treewidth of $H$ in \cref{kPowerBasic} is improved from $O(k^{12})$ to $O(k^7)$ in \cref{Powers}.

@@

\cref{Examples} presents several examples of graph classes that can be obtained from a shortcut system applied to a planar graphs, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of the above results are applicable to these examples.

All of the above results apply, where instead of planar graphs, we consider graphs of bounded Euler genus.

@@

This section defines concepts from from the literature that will be important for our work.

@@

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has bounded treewidth. They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $t$.

@@

Apex-minor-free graphs are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

\section{Our Results}

The following result is the main theorem of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so to does $G^\PP$ for every $(k,d)$-shortcut system $\PP$.

@@

The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

@@

%%%%%%%%%%%%%%%%
\subsection{Prior Work}
\label{PriorWork}

@@

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, suppose that an $n$-vertex graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$. \citet{dujmovic.joret.ea:planar} proved that $G$ has layered treewidth at most $(k+1)\ell$. So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.

>>>>

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
% Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to this setting, and are the first results of this type for non-proper-minor-closed graph classes.

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).

@@

We actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

\cref{PlanarBasic,ShortcutProduct,AddDummy} immediately imply \cref{kPlanarBasic} with $O(k^{11})$ instead of $O(k^5)$. Some further observations reduce the bound to $O(k^5)$; see \cref{Tools}.

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, we use $\dist_G(v,w)$ to denote the length of a shortest path with endpoints $v$ and $w$.  If $v$ and $w$ are in different components of $G$ then we use the convention that $\dist_G(v,w)=\infty$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

% Note that the dependence on $k$ in the bound on the treewidth of $H$ in \cref{kPowerBasic} is improved from $O(k^{12})$ to $O(k^7)$ in \cref{Powers}. 
% PM: I removed this because the bound is exponential in $k$ anyway. You only see the difference when you separate treewidth and layered width.

@@

\cref{Examples} presents several examples of graph classes that can be obtained from a shortcut system applied to a planar graph, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of these results also apply, where instead of planar graphs, we consider graphs of bounded Euler genus.   All of the applications in \cref{Applications}  are applicable to these examples.

@@

This section defines concepts and results from the literature that will be important for our work.

@@

\subsection{Previous Work on Partitions of Minor-Closed Classes}
\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has bounded treewidth. They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $t$.

@@

\section{New Results}

Apex-minor-free graphs, addressed by \cref{ApexMinorFree}, are the largest minor-closed class that admit bounded layered partitions~\citep{dujmovic.joret.ea:planar}. However, the family of $k$-planar graphs is not minor-closed.  A graph $G'$ obtained from a $k$-planar graph $G$ by edge deletions and edge contractions may or may not be $k$-planar. Indeed, \citet{dujmovic.eppstein.ea:structure} construct 1-planar graphs that contain arbitrarily large complete graph minors. Our results for $k$-planar graphs are the first of this type for a non-minor-closed class.

The following result is the main theorem of the paper. Loosely speaking, it shows that if a graph $G$ admits bounded layered partitions, then so to does $G^\PP$ for every $(k,d)$-shortcut system $\PP$ of $G$.

@@

The relationship between $k$-planar graphs and $(k+1,2)$-shortcut systems of planar graphs shows that \cref{modern-major-general} has immediate applications to $k$-planar graphs.  The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

@@

%%%%%%%%%%%%%%%%
\subsection{Previous Work on the Structure of $(g,k)$-Planar Graphs}
\label{PriorWork}

@@

If a graph class admits bounded layered partitions, then it also has bounded layered treewidth. In particular, suppose that an $n$-vertex graph $G$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $k$. \citet{dujmovic.joret.ea:planar} proved that $G$ has layered treewidth at most $(k+1)\ell$. 
% So any property that holds for graphs of bounded layered treewidth also holds for $G$. For example, Norin proved that every $n$-vertex graph with layered treewidth at most $t$ has treewidth less than $2\sqrt{tn}$ (see \citep{dujmovic.morin.ea:layered}). Thus, $G$ has treewidth at most $2\sqrt{(k+1)\ell n}$. This in turn leads to a $O(\sqrt{(k+1)\ell n})$ balanced separator theorem for such graphs \citep{RS-II}. So, in the case of $k$-planar graphs, \cref{k-planar} implies an analogue of the separator theorem of \citet{FP08} mentioned in the introduction, albeit with worse dependence on $k$. 
What sets layered partitions apart from layered treewidth is that they lead to constant upper bounds on the queue-number  and non-repetitive chromatic number, whereas for both these parameters, the best known upper bound obtainable via layered treewidth is $O(\log n)$; see \cref{Applications}.
